---
title: "PM566 Final Project"
author: "**Yating Zeng**"
date: "`r Sys.Date()`"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
        keep_md : TRUE 
    github_document:
      html_preview: false
always_allow_html: TRUE
---
This is my PM566 Final Project website. 
https://yating-zeng.github.io/PM566_Lab12/Final_Project.html
and Supplementary website.
https://yating-zeng.github.io/PM566_Lab12/Final_Project_Supplementary.html
<br>


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, echo = FALSE, warning = FALSE, message = FALSE)
```

# Introduction

COVID-19 has been here for around 3 years, with vaccine widely used. It would be likely that some of the people tend to not take the vaccine than the others. Thus, in this project, the question of my interest is: What is the association between age and the vaccination status (at least one dose & completed a primary series) in California state? 
For this project, I’ll use the dataset on Covid-19 vaccination from the Centers for Disease Control and Prevention (CDC) website, which provided data for select demographic characteristics (age, sex, and age by sex) of people receiving COVID-19 vaccinations in the United States at the national and jurisdictional levels, fitting my analysis interest well. All the data were cumulative data, which were counted since the date it started observing.

<br>

# Methods

## 1.Dataset 

In this project, the dataset used was a public resource from CDC website, named “COVID-19 Vaccination Age and Sex Trends in the United States, National and Jurisdictional”. The link of the dataset is shown below: https://data.cdc.gov/Vaccinations/COVID-19-Vaccination-Age-and-Sex-Trends-in-the-Uni/5i5k-6cmh. The CSV file of the data was then downloaded and read into R studio for further analysis in this project.

```{r }
library(readr)
library(tidyverse)
library(dplyr)
library(stringr)
if(!require(kableExtra)) install.packages("kableExtra")
library(kableExtra)
if(!require(plotly)) install.packages("plotly", repos = "http://cran.us.r-project.org")
library(plotly)
if(!require(plotly)) install.packages("DT")
library(DT)
```

```{r read-data}
#read in the dataset

if (!file.exists("COVID19_Vaccination.csv")){
  library("RSocrata")
  vaccination <- read.socrata(
                 "https://data.cdc.gov/resource/n8mc-b4w4.json",
                  app_token = "KS8vICWuRMDR6QzLnGP7SVO1a",
                  email     = "yatingzeng18@gmail.com",
                  password  = "Ttzyt119089838--"
  )
}

vaccination <- read_csv("COVID19_Vaccination.csv")
```

<br>

## 2.Data cleaning, wrangling and EDA

After checking the summary of the content of the dataset, the dimensions and the original properties for each variable were known. I filtered the data to create a new dataset to keep only the information of California. For simplifying the typing in analysis, 7 columns were renamed to be shorter. Then the proportion of missing values of each column and column "Demographic_Category" were checked. Considering that age and vaccination status of primary dose series were the main factors towards this analysis, the information about "Booster",  "Age_unknown" and all the "Age>65" levels of the "Demographic_Category" column, and and the missing values of "dose1"(count of people take at least one dose) and "census" (census statistics used for calculating the percentage of vaccination) were removed.

Because the information this dataset was about the information strongly rely on time series, and all the statistics were cumulative data, a new variable "date" was created for further reorder the data by the time recorded.
Based on the category from "Demographic_Category" variable (now named "cat"), the original dataset was split into 4 subset for better analysis, which are 1. objects from both sex categorized only by age level; 2. objects were all females categorized by age level; 3. objects were all males categorized by age level; and 4. objects from both sex categorized only by sex level. 

Totally 25 tables and figures were created, with 1 table and 4 figures would shown below and the others to be the supplementary. All the tables and figure were planed to create by 2 vaccination status ("at least one dose" and "completed a primary series") and 4 categorical groups ("age"; "female_age"; "male_age"; and "sex"). The summary tables and figures would show the minimum, 1st quantile, median, 3nd quantile, maximum, and the number of recorded objects of "the percentage of people" with the 2 kinds of vaccination status grouped by age, sex or age groups stratified by sex. The reason to use data stratified by sex was to remove the possible confounding effect from sex on the association between vaccination status and age level. The association between the age and vaccination status would be shown with scatterplots.

```{r keep data of CA}

#select only the data of CA
ca_vac <- vaccination[which(vaccination$Location == "CA"), ]
#str(ca_vac)

#reorder the dataset by Demographic_Category and then date
ca_vac <- ca_vac[order( ca_vac[,3], ca_vac[,1] ),]
```

```{r Change the names of the key variables}
#head(ca_vac)
#simplified the variable names
colnames(ca_vac)[3]  <- "cat"
colnames(ca_vac)[5]  <- "dose1"
colnames(ca_vac)[6]  <- "series"
colnames(ca_vac)[9]  <- "dose1_pct"
colnames(ca_vac)[10] <- "series_pct"
colnames(ca_vac)[11] <- "booster_pct"
colnames(ca_vac)[12] <- "secbooster_pct"
```

```{r remove missing values}
#checking the proportion of missing values
#(colMeans(is.na(ca_vac)))*100

vac <- ca_vac[which(ca_vac$cat != "Age_Unknown"), ] 
vac <- vac %>%
  filter(!is.na(vac$dose1),!is.na(vac$census))

#check the missing value again
#(colMeans(is.na(vac)))*100

#learn more about the dataset
#check about the "Demographic_Category", "Dose1_pct",and "series_pct"
#unique(vac$cat)
#summary(vac$dose1_pct)
#summary(vac$series_pct)
```

```{r}
#create new variables about date
vac$Date  <- substr(vac$Date, 0, 10)
vac$year  <- substr(vac$Date, 7, 10)
vac$month <- substr(vac$Date, 0, 2)
vac$day   <- substr(vac$Date, 4, 5)

#sort the data by date 
vac1 <- vac[with(vac, order(year, month, day)), ]

#create a new "date" numeric variable with the time order acceptable for reoder dataset
vac1 <- mutate(vac1, date = paste(year, month, day))
vac1$date <- str_replace_all(vac1$date, fixed(" "), "")

vac1$year  <- as.numeric(vac1$year)
vac1$month <- as.numeric(vac1$month)
vac1$day   <- as.numeric(vac1$day)

```

```{r remove the informarion}
#remove the some information of no interest: booster information; the data of level "ages 65+"
vac1 = subset(vac1, select = -c(Booster_Doses, Second_Booster, booster_pct, secbooster_pct) )
vac1 <- vac1 %>%
  filter(vac1$cat != "Ages_65+_yrs",
               vac1$cat != "Female_Ages_65+_yrs",
               vac1$cat != "Male_Ages_65+_yrs"
               )

#find that there is a unreasonable order for the level 5-11

#rename the level of 5-11 to 05-11
vac1$cat <- str_replace_all(vac1$cat, fixed("Female_Ages_5-11_yrs"), "Female_Ages_05-11_yrs")
vac1$cat <- str_replace_all(vac1$cat, fixed("Male_Ages_5-11_yrs"),   "Male_Ages_05-11_yrs")
vac1$cat <- str_replace_all(vac1$cat, fixed("Ages_5-11_yrs"),        "Ages_05-11_yrs")
vac1$cat <- str_replace_all(vac1$cat, fixed("Female_Ages_2-4_yrs"),  "Female_Ages_02-04_yrs")
vac1$cat <- str_replace_all(vac1$cat, fixed("Male_Ages_2-4_yrs"),    "Male_Ages_02-04_yrs")
vac1$cat <- str_replace_all(vac1$cat, fixed("Ages_2-4_yrs"),         "Ages_02-04_yrs")

```

```{r}
#splitting the data by "cat" level into 4 subset: "age"; "Female_age"; "Male_age"; "sex"
vac1$CAT = substr(vac1$cat, 0, 1)
#build a subset for 
vac_age<- vac1 %>%
  filter(vac1$CAT == "A")
vac_Fage<- vac1 %>%
  filter(vac1$CAT == "F")
vac_Mage<- vac1 %>%
  filter(vac1$CAT == "M")
vac_sex<- vac1 %>%
  filter(vac1$CAT == "S")
```

<br>

# Results

Considering that all the results for people who completed a series dose (Supplementary table 4-7, 11-15,17-20) were consistent with the results for people with at least one dose, thus, in this report, we would focus on the showing the results of analysis for people with at least one dose, which contained more information.

## 1. Summary tables

From the summary table of the percent of people with at least one dose grouped by age (Table 1), we could notice that excepting a little part of the data, most part the data were showing a trend that the statistics (the minimum, 1st quantile, median, 3nd quantile, maximum) of Percent of people with at least one dose would be larger when the age level was higher. And for the major part of observed objects, who were aged from 12-17 years old to 75+ years old, the final vaccination rate will went up to around 90%, while for the low age-level(<5 years) objects, the vaccination rate were all under or around 10%, and for 5-11 years objects the vaccination rate stayed in the middle, which is around 50%. This trend were consistent with the results form all the other 2 summary tables of Percent of people with at least one dose grouped by age stratified by sex (Supplementary table 1 & 2).



```{r summary table for dose1}
# summary table for dose1
vac_age_tab <- vac_age %>% group_by(cat) %>%
                   summarise(
                     min    = min(dose1_pct, na.rm = T),
                     q1     = quantile(dose1_pct, 0.25),
                     median = median(dose1_pct),
                     q3     = quantile(dose1_pct, 0.75),
                     max    = max(dose1_pct, na.rm = T), 
                     days_recorded    = sum(!is.na(dose1_pct))
                     ) %>%  arrange(cat)

vac_Fage_tab <- vac_Fage %>% group_by(cat) %>%
                   summarise(
                     min    = min(dose1_pct, na.rm = T),
                     q1     = quantile(dose1_pct, 0.25),
                     median = median(dose1_pct),
                     q3     = quantile(dose1_pct, 0.75),
                     max    = max(dose1_pct, na.rm = T), 
                     days_recorded    = sum(!is.na(dose1_pct))
                     ) %>%  arrange(cat)

vac_Mage_tab <- vac_Mage %>% group_by(cat) %>%
                   summarise(
                     min           = min(dose1_pct, na.rm = T),
                     q1            = quantile(dose1_pct, 0.25),
                     median        = median(dose1_pct),
                     q3            = quantile(dose1_pct, 0.75),
                     max           = max(dose1_pct, na.rm = T), 
                     days_recorded  = sum(!is.na(dose1_pct))
                     ) %>%  arrange(cat)

vac_sex_tab <- vac_sex %>% group_by(cat) %>%
                   summarise(
                     min    = min(dose1_pct, na.rm = T),
                     q1     = quantile(dose1_pct, 0.25),
                     median = median(dose1_pct),
                     q3     = quantile(dose1_pct, 0.75),
                     max    = max(dose1_pct, na.rm = T), 
                     days_recorded    = sum(!is.na(dose1_pct))
                     ) %>%  arrange(cat)
```
<br>

**Table 1.Summary of Percent of people with at least one dose grouped by age**
```{r}
datatable(vac_age_tab)
```


```{r summary table for series}
# summary table for series dose
vac_age_tab2 <- vac_age %>% group_by(cat) %>%
                   summarise(
                     min    = min(series_pct, na.rm = T),
                     q1     = quantile(series_pct, 0.25, na.rm = T),
                     median = median(series_pct, na.rm = T),
                     q3     = quantile(series_pct, 0.75, na.rm = T),
                     max    = max(series_pct, na.rm = T), 
                     days_recorded    = sum(!is.na(series_pct))
                     ) %>%  arrange(cat) 

vac_Fage_tab2 <- vac_Fage %>% group_by(cat) %>%
                   summarise(
                     min    = min(series_pct, na.rm = T),
                     q1     = quantile(series_pct, 0.25, na.rm = T),
                     median = median(series_pct, na.rm = T),
                     q3     = quantile(series_pct, 0.75, na.rm = T),
                     max    = max(series_pct, na.rm = T), 
                     days_recorded    = sum(!is.na(series_pct))
                     ) %>%  arrange(cat)

vac_Mage_tab2 <- vac_Mage %>% group_by(cat) %>%
                   summarise(
                     min    = min(series_pct, na.rm = T),
                     q1     = quantile(series_pct, 0.25, na.rm = T),
                     median = median(series_pct, na.rm = T),
                     q3     = quantile(series_pct, 0.75, na.rm = T),
                     max    = max(series_pct, na.rm = T), 
                     days_recorded    = sum(!is.na(series_pct))
                     ) %>%  arrange(cat)

vac_sex_tab2 <- vac_sex %>% group_by(cat) %>%
                   summarise(
                     min    = min(series_pct, na.rm = T),
                     q1     = quantile(series_pct, 0.25, na.rm = T),
                     median = median(series_pct, na.rm = T),
                     q3     = quantile(series_pct, 0.75, na.rm = T),
                     max    = max(series_pct, na.rm = T), 
                     days_recorded    = sum(!is.na(series_pct))
                     ) %>%  arrange(cat)
```
<br>

## 2. Summary figures

The summary figures (Figure 1) showed the similar trend mentioned above relatively. Excepting those information, we still could notice that for the object age level from 5-11 years old to 65-74 years old, with the age level went up, the major part of the statistics would with higher values, which means it might take shorter time to have a relatively high vaccination rate for those people with higher age level. 

```{r summary graphs for dose1, fig.height = 5, fig.width = 8}
#summary graphs for dose1
age1 <- vac_age %>%
    ggplot(aes(x=date, y=dose1_pct)) +
    geom_boxplot(mapping = aes(x = cat, y = dose1_pct, fill = cat)) +
    theme(axis.text.x  = element_text(angle = 60, hjust = 1),
          plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
    labs(x = "Age group", y = "Percent of people with at least one dose", fill = 'Age group',
         title = "Figure 1.Percent of people with at least one dose grouped by age")
ggplotly(age1)

#Fage1 <- vac_Fage %>%
#    ggplot(aes(x=date, y=dose1_pct)) +
#    geom_boxplot(mapping = aes(x = cat, y = dose1_pct, fill = cat)) +
#    theme(axis.text.x = element_text(angle = 60, hjust = 1),
#          plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
#    labs(x = "Age group", y = "Percent of people with at least one dose", fill = 'Age group',
#         title = "Spplementary figure 8.Percent of people with at least one dose of females grouped by age") 


#Mage1 <- vac_Mage %>%
#    ggplot(aes(x=date, y=dose1_pct)) +
#    geom_boxplot(mapping = aes(x = cat, y = dose1_pct, fill = cat)) +
#    theme(axis.text.x = element_text(angle = 60, hjust = 1),
#          plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
#    labs(x = "Age group", y = "Percent of people with at least one dose", fill = 'Age group',
#         title = "Spplementary figure 9.Percent of people with at least one dose of males grouped by age")


#sex1 <- vac_sex %>%
#    ggplot(aes(x=date, y=dose1_pct)) +
#    geom_boxplot(mapping = aes(x = cat, y = dose1_pct, fill = cat)) +
#    theme(plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
#    labs(x = "Sex", y = "Percent of people with at least one dose", fill='Sex group',
#         title = "Spplementary figure 10.Percent of people with at least one dose grouped by sex")
```

From the line plot (figures 2) suggests that there would be difference of the percent of having at least one dose between female and male, making it to be reasonable to use the stratified data for analyze the association between age and vaccination rate.

```{r}
sex01 <- vac_sex %>%
    ggplot(aes(x = date, y = dose1_pct)) + 
    geom_point(mapping = aes(x = date, y = dose1_pct, color = cat)) +
    scale_x_discrete(breaks=
          c("20201213","20210601","20211201","20220601","20221019")) +
    theme(plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
    labs(x = "Date(yyyymmdd)", y = "Percent of people with at least one dose", col="Sex group", 
         title = "figure 2.Percent of people completed a primary seriese grouped by sex")
ggplotly(sex01)
```


```{r summary graphs for series doses, fig.height = 5, fig.width = 9}
#summary graphs for series doses
#age2 <- vac_age %>%
#    ggplot(aes(x=date, y=series_pct)) +
#    geom_boxplot(mapping = aes(x = cat, y = dose1_pct, fill = cat)) +
#    theme(axis.text.x = element_text(angle = 60, hjust = 1),
#          plot.caption = element_text(hjust = 0.5, size=rel(1.2))) +
#    labs(x = "Age group", y = "Percent of people completed a primary series", fill = 'Age group',
#         title = "Spplementary figure 11.Percent of people completed a primary series grouped by age")


#Fage2 <- vac_Fage %>%
#    ggplot(aes(x=date, y=series_pct)) +
#    geom_boxplot(mapping = aes(x = cat, y = dose1_pct, fill = cat)) +
#    theme(axis.text.x = element_text(angle = 60, hjust = 1),
#          plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
#    labs(x = "Age group", y = "Percent of people completed a primary series", fill = 'Age group',
#         title = "Spplementary figure 12.Percent of people completed a primary series of males grouped by age")

#Mage2 <- vac_Mage %>%
#    ggplot(aes(x=date, y=series_pct)) +
#    geom_boxplot(mapping = aes(x = cat, y = dose1_pct, fill = cat)) +
#    theme(axis.text.x = element_text(angle = 60, hjust = 1),
#          plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
#    labs(x = "Age group", y = "Percent of people completed a primary series", fill = 'Age group',
#         title = "Spplementary figure 13.Percent of people completed a primary series of males grouped by age")

#sex2 <- vac_sex %>%
#    ggplot(aes(x=date, y=series_pct)) +
#    geom_boxplot(mapping = aes(x = cat, y = dose1_pct, fill = cat)) +
#    theme(plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
#    labs(x = "Sex", y = "Percent of people completed a primary seriese", fill = 'Sex group',
#         title = "Spplementary figure 14.Percent of people completed a primary seriese grouped by sex in boxplot")

#sex02 <- vac_sex %>%
#    ggplot(aes(x = date, y = dose1_pct)) + 
#    geom_point(mapping = aes(x = date, y = dose1_pct, color = cat)) +
#    scale_x_discrete(breaks=
#          c("20201213","20210601","20211201","20220601","20221019")) +
#    theme(plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
#    labs(x = "Date(yyyymmdd)", y = "Percent of people with at least one dose", col="Sex group", 
#         title = "Spplementary figure 15.Percent of people completed a primary seriese grouped by sex in line plot")
```

<br>

## 3. Visualization of the association

These 2 figures (Figure 3-4) all verified the trend that after adjusting by sex, the vaccination rate would be higher with the age level being higher for the same time point, and the objects with higher age might take shorter time to have a relatively high vaccination rate. Which needs to be mentioned is that, this trend was also observed in the low age-level group(<5 years), but with obviously lower vaccination rate than the major part of the sample objects.

```{r visualization for first dose}
#visualization for first dose
#age3 <- vac_age %>%
#    ggplot(aes(x = date, y = dose1_pct)) + 
#    geom_point(mapping = aes(x = date, y = dose1_pct, color = cat)) +
#    scale_x_discrete(breaks=
#          c("20201213","20210601","20211201","20220601","20221019")) +
#    theme(plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
#    labs(x = "Date(yyyymmdd)", y = "Percent of people with at least one dose", col="Age group",
#         title = "Spplementary figure 16.2020-2022 Percent of people with at least one dose grouped by age")

Fage03 <- vac_Fage %>%
    ggplot(aes(x = date, y = dose1_pct)) + 
    geom_point(mapping = aes(x = date, y = dose1_pct, color = cat)) +
    scale_x_discrete(breaks=
          c("20201213","20210601","20211201","20220601","20221019")) +
    theme(plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
    labs(x = "Date(yyyymmdd)", y = "Percent of people with at least one dose", col="Age group", 
         title = "Figure 3.Percent of people with at least one dose of females grouped by age")
ggplotly(Fage03)

Mage03 <- vac_Mage %>%
    ggplot(aes(x=date, y=dose1_pct)) + 
    geom_point(mapping = aes(x = date, y = dose1_pct, color = cat)) +
    theme(plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
    scale_x_discrete(breaks=
          c("20201213","20210601","20211201","20220601","20221019")) +
    labs(x = "Date(yyyymmdd)", y = "Percent of people with at least one dose", col="Age group", 
         title = "Figure 4.Percent of people with at least one dose of males grouped by age") +
    scale_fill_discrete(name = "Age group")
ggplotly(Mage03)
```

```{r visualization for series dose, fig.height = 6, fig.width = 10}
#visualization for series dose
#age4 <- vac_age %>%
#    ggplot(aes(x=date, y=series_pct)) + 
#    geom_point(mapping = aes(x = date, y = dose1_pct, color = cat))  +
#    theme(plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
#    scale_x_discrete(breaks=          c("20201213","20210601","20211201","20220601","20221019")) +
#    labs(x = "Date(yyyymmdd)", y = "Percent of people completed a primary series", col="Age group",
#         title = "Spplementary figure 17.Percent of people completed a primary series grouped by age")
 
#sex4 <- vac_sex %>%
#    ggplot(aes(x=date, y=series_pct)) + 
#    geom_point(mapping = aes(x = date, y = dose1_pct, color = cat))  +
#    theme(plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
#    scale_x_discrete(breaks=
#          c("20201213","20210601","20211201","20220601","20221019")) +
#    labs(x = "Date(yyyymmdd)", y = "Percent of people completed a primary series", col="Sex group",
#         title = "Spplementary figure 18.Percent of people completed a primary series grouped by sex")

#Fage4 <- vac_Fage %>%
#    ggplot(aes(x=date, y=series_pct)) + 
#    geom_point(mapping = aes(x = date, y = dose1_pct, color = cat))  +
#    theme(plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
#    scale_x_discrete(breaks=
#          c("20201213","20210601","20211201","20220601","20221019")) +
#    labs(x = "Date(yyyymmdd)", y = "Percent of people completed a primary series", col="Age group",
#         title = "Spplementary figure 19.Percent of people completed a primary series of females grouped by age")

#Mage4 <- vac_Mage %>%
#    ggplot(aes(x=date, y=series_pct)) + 
#    geom_point(mapping = aes(x = date, y = dose1_pct, color = cat))  +
#    theme(plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
#    scale_x_discrete(breaks=
#          c("20201213","20210601","20211201","20220601","20221019")) +
#    labs(x = "Date(yyyymmdd)", y = "Percent of people completed a primary series", col="Age group",
#         caption = "Spplementary figure 20.Percent of people completed a primary series of males grouped by age") +
#    guides(fill=guide_legend(title="Age group"))
```

<br>

# Conclusion

We could believe that there could be an association between age and the two vaccination status (at least one dose & completed a primary series) in California state.For both 2 kinds of vaccination status(take at least one dose & with completed series) and both sex, the vaccination rate would be higher with the age level being higher for the same time point, and the objects with higher age might also take shorter time to have a relatively high vaccination rate. And the final vaccination rate would be higher with the age level being higher, but the rate for people with age less than 5 years old would keep in a low level, even though they follow the same trend mentioned above.

<br>