---
title: "PM566 Final Project"
author: '**Yating Zeng**'
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
  github_document:
    html_preview: no
  html_document:
    toc: yes
    toc_float: yes
    keep_md: yes
always_allow_html: yes
---
Link to my report PDF: [**Report**](https://github.com/yating-zeng/PM566_Final_Project/blob/main/Report.pdf)       [[download]](https://github.com/yating-zeng/PM566_Final_Project/raw/main/Report.pdf)

Link to my supplements PDF: [**Supplements**](https://github.com/yating-zeng/PM566_Final_Project/blob/main/Supplements.pdf) [[download]](https://github.com/yating-zeng/PM566_Final_Project/raw/main/Supplements.pdf)

<br>


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, echo = FALSE, warning = FALSE, message = FALSE)
```

# Introduction

COVID-19 has been here for around 3 years, with vaccine widely used. It would be likely that some of the people tend to not take the vaccine than the others. Thus, in this project, the question of my interest is: What is **the association between age and the COVID-19 vaccination status** (at least one dose & completed a primary series) in California state? 
For this project, I’ll use the dataset on Covid-19 vaccination from the Centers for Disease Control and Prevention (CDC) website, which provided data for select demographic characteristics (age, sex, and age by sex) of people receiving COVID-19 vaccinations in the United States at the national and jurisdictional levels, fitting my analysis interest well. All the data were cumulative data, which were counted since the date it started observing.

<br>

# Methods

## 1.Dataset 

In this project, the dataset used was a public resource from *CDC website*, named “COVID-19 Vaccination Age and Sex Trends in the United States, National and Jurisdictional”. The link of the dataset is shown below: https://data.cdc.gov/Vaccinations/COVID-19-Vaccination-Age-and-Sex-Trends-in-the-Uni/5i5k-6cmh. The CSV file of the data was then downloaded and read into R studio for further analysis in this project.

```{r }
library(readr)
library(tidyverse)
library(dplyr)
library(stringr)
if(!require(kableExtra)) install.packages("kableExtra")
library(kableExtra)
if(!require(plotly)) install.packages("plotly", repos = "http://cran.us.r-project.org")
library(plotly)
if(!require(plotly)) install.packages("DT")
library(DT)
```

```{r read-data}
#read in the dataset

if (!file.exists("COVID19_Vaccination.csv")){
  library("RSocrata")
  vaccination <- read.socrata(
                 "https://data.cdc.gov/resource/n8mc-b4w4.json",
                  app_token = "KS8vICWuRMDR6QzLnGP7SVO1a",
                  email     = "yatingzeng18@gmail.com",
                  password  = "Ttzyt119089838--"
  )
}

vaccination <- read_csv("COVID19_Vaccination.csv")
```

<br>

## 2.Data cleaning, wrangling and EDA

After checking the summary of the content of the dataset, the dimensions and the original properties for each variable were known. I filtered the data to create a new dataset to keep only the information of California. Totally *25 tables and figures* were created, with 5 figures would shown below and more information could be found in the supplementary. All the tables and figure were planed to create by 2 vaccination status ("at least one dose" and "completed a primary series") and 4 categorical groups ("age"; "female_age"; "male_age"; and "sex"). The summary tables and figures would show the minimum, 1st quantile, median, 3nd quantile, maximum, and the number of recorded objects of "the percentage of people" with the 2 kinds of vaccination status grouped by age, sex or age groups stratified by sex. The association between the age and vaccination status would be shown with scatterplots.

```{r keep data of CA}

#select only the data of CA
ca_vac <- vaccination[which(vaccination$Location == "CA"), ]
#str(ca_vac)

#reorder the dataset by Demographic_Category and then date
ca_vac <- ca_vac[order( ca_vac[,3], ca_vac[,1] ),]
```

```{r Change the names of the key variables}
#head(ca_vac)
#simplified the variable names
colnames(ca_vac)[3]  <- "cat"
colnames(ca_vac)[5]  <- "dose1"
colnames(ca_vac)[6]  <- "series"
colnames(ca_vac)[9]  <- "dose1_pct"
colnames(ca_vac)[10] <- "series_pct"
colnames(ca_vac)[11] <- "booster_pct"
colnames(ca_vac)[12] <- "secbooster_pct"
```

```{r remove missing values}
#checking the proportion of missing values
#(colMeans(is.na(ca_vac)))*100

vac <- ca_vac[which(ca_vac$cat != "Age_Unknown"), ] 
vac <- vac %>%
  filter(!is.na(vac$dose1),!is.na(vac$census))

#check the missing value again
#(colMeans(is.na(vac)))*100

#learn more about the dataset
#check about the "Demographic_Category", "Dose1_pct",and "series_pct"
#unique(vac$cat)
#summary(vac$dose1_pct)
#summary(vac$series_pct)
```

```{r}
#create new variables about date
vac$Date  <- substr(vac$Date, 0, 10)
vac$year  <- substr(vac$Date, 7, 10)
vac$month <- substr(vac$Date, 0, 2)
vac$day   <- substr(vac$Date, 4, 5)

#sort the data by date 
vac1 <- vac[with(vac, order(year, month, day)), ]

#create a new "date" numeric variable with the time order acceptable for reoder dataset
vac1 <- mutate(vac1, date = paste(year, month, day))
vac1$date <- str_replace_all(vac1$date, fixed(" "), "")

vac1$year  <- as.numeric(vac1$year)
vac1$month <- as.numeric(vac1$month)
vac1$day   <- as.numeric(vac1$day)

```

```{r remove the informarion}
#remove the some information of no interest: booster information; the data of level "ages 65+"
vac1 = subset(vac1, select = -c(Booster_Doses, Second_Booster, booster_pct, secbooster_pct) )
vac1 <- vac1 %>%
  filter(vac1$cat != "Ages_65+_yrs",
               vac1$cat != "Female_Ages_65+_yrs",
               vac1$cat != "Male_Ages_65+_yrs"
               )

#find that there is a unreasonable order for the level 5-11

#rename the level of 5-11 to 05-11
vac1$cat <- str_replace_all(vac1$cat, fixed("Female_Ages_5-11_yrs"), "Female_Ages_05-11_yrs")
vac1$cat <- str_replace_all(vac1$cat, fixed("Male_Ages_5-11_yrs"),   "Male_Ages_05-11_yrs")
vac1$cat <- str_replace_all(vac1$cat, fixed("Ages_5-11_yrs"),        "Ages_05-11_yrs")
vac1$cat <- str_replace_all(vac1$cat, fixed("Female_Ages_2-4_yrs"),  "Female_Ages_02-04_yrs")
vac1$cat <- str_replace_all(vac1$cat, fixed("Male_Ages_2-4_yrs"),    "Male_Ages_02-04_yrs")
vac1$cat <- str_replace_all(vac1$cat, fixed("Ages_2-4_yrs"),         "Ages_02-04_yrs")

```

```{r}
#splitting the data by "cat" level into 4 subset: "age"; "Female_age"; "Male_age"; "sex"
vac1$CAT = substr(vac1$cat, 0, 1)
#build a subset for 
vac_age<- vac1 %>%
  filter(vac1$CAT == "A")
vac_Fage<- vac1 %>%
  filter(vac1$CAT == "F")
vac_Mage<- vac1 %>%
  filter(vac1$CAT == "M")
vac_sex<- vac1 %>%
  filter(vac1$CAT == "S")
```

<br>

# Results

## **1. Vaccination status trends of "At least one dose" and "A primary series" with little difference**

From the summary figures shown bellow, we could notice that both the percent of people with at least one dose grouped by age (Figure 1) and the percent of people completed a primary series grouped by age (Figure 2) are showing consistent trends and similar data structure. Excepting a little part of the data, most of the data were showing a trend that the statistics (the minimum, 1st quantile, median, 3nd quantile, maximum) of vaccination rate would increase, when the age level was higher. Thus, we would focus on the the results of the sample with at least one does for the next in this report.

```{r summary figures, fig.height = 5, fig.width = 8}
#summary graphs for dose1
age1 <- vac_age %>%
    ggplot(aes(x=date, y=dose1_pct)) +
    geom_boxplot(mapping = aes(x = cat, y = dose1_pct, fill = cat)) +
    theme(axis.text.x  = element_text(angle = 60, hjust = 1),
          plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
    labs(x = "Age group", y = "Percent of people with at least one dose", fill = 'Age group',
         title = "Figure 1.Percent of people with at least one dose grouped by age")
#summary graphs for series doses
age2 <- vac_age %>%
    ggplot(aes(x=date, y=series_pct)) +
    geom_boxplot(mapping = aes(x = cat, y = dose1_pct, fill = cat)) +
    theme(axis.text.x = element_text(angle = 60, hjust = 1),
          plot.caption = element_text(hjust = 0.5, size=rel(1.2))) +
    labs(x = "Age group", y = "Percent of people completed a primary series", fill = 'Age group',
         title = "Figure 2.Percent of people completed a primary series grouped by age")
```
### {.tabset}

#### Figure 1.

```{r}
ggplotly(age1)
```

#### Figure 2.

```{r}
ggplotly(age2)
```

### {-}

<br>

## **2. Females have higher vaccination rate than males**

The results for the percent group by sex shown in Figure 1. suggests that compared to the male, females would have higher vaccination rate across all the time recorded, and went up to a higher level in the end. Considering over this issue, we would better to analyze the data stratified by gender to prevent possible confounding effect.

```{r}
sex01 <- vac_sex %>%
    ggplot(aes(x = date, y = dose1_pct)) + 
    geom_point(mapping = aes(x = date, y = dose1_pct, color = cat)) +
    scale_x_discrete(breaks=
          c("20201213","20210601","20211201","20220601","20221019"))+
    scale_color_discrete(labels = c("Female", "Male")) +
    theme(plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
    labs(x = "Date(yyyymmdd)", y = "Percent of people with at least one dose", col="Sex group", 
         title = "Figure 3.Percent of people completed a primary seriese grouped by sex")
ggplotly(sex01)
```

## **3. The vaccination rate would be higher with the age level increased**

Based on the figures bellow (Figure 5 & 6), we could find that in both gender, the vaccination rate were higher with the age level being higher for the same time point, and the objects with higher age might take shorter time to have a relatively high vaccination rate. Which needs to be mentioned is that, this trend was also observed in the low age-level group(<5 years), but with obviously lower vaccination rate than the major part of the sample objects.


```{r visualization for first dose}
Fage03 <- vac_Fage %>%
    ggplot(aes(x = date, y = dose1_pct)) + 
    geom_point(mapping = aes(x = date, y = dose1_pct, color = cat)) +
    scale_x_discrete(breaks=
          c("20201213","20210601","20211201","20220601","20221019")) +
    theme(plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
    labs(x = "Date(yyyymmdd)", y = "Percent of people with at least one dose of female", col="Age group", 
         title = "Figure 4.Percent of people with at least one dose of females") +
    scale_fill_discrete(name = "Age group")


Mage03 <- vac_Mage %>%
    ggplot(aes(x=date, y=dose1_pct)) + 
    geom_point(mapping = aes(x = date, y = dose1_pct, color = cat)) +
    theme(plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
    scale_x_discrete(breaks=
          c("20201213","20210601","20211201","20220601","20221019")) +
    labs(x = "Date(yyyymmdd)", y = "Percent of people with at least one dose", col="Age group", 
         title = "Figure 5.Percent of people with at least one dose of males") +
    scale_fill_discrete(name = "Age group")

```

### {.tabset}

#### Figure 4.

```{r}
ggplotly(Fage03)
```

#### Figure 5.

```{r}
ggplotly(Mage03)
```

### {-}

# Conclusion

We could believe that there could be an association between age and the two vaccination status (at least one dose & completed a primary series) in California state. For both two kinds of vaccination status(take at least one dose & with completed series) and both sex, the vaccination rate would be higher with the age level being higher for the same time point, and the objects with higher age might also take shorter time to have a relatively high vaccination rate. And the final vaccination rate would be higher with the age level being higher, but the rate for people with age less than 5 years old would keep in a low level, even though they follow the same trend mentioned above.

<br>