---
title: "PM566 Final Project Spplementary"
author: "**Yating Zeng**"
date: "`r Sys.Date()`"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
        keep_md : TRUE 
    github_document:
      html_preview: false
always_allow_html: TRUE
---
Link to my report pdf: [Report](https://github.com/yating-zeng/PM566_Final_Project/blob/main/Report.pdf)       [[download]](https://github.com/yating-zeng/PM566_Final_Project/raw/main/Report.pdf)

Link to this Supplement pdf: [Supplements](https://github.com/yating-zeng/PM566_Final_Project/blob/main/Supplements.pdf) [[download]](https://github.com/yating-zeng/PM566_Final_Project/raw/main/Supplements.pdf)
<br>


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, echo = FALSE, warning = FALSE, message = FALSE)
```

```{r }
library(readr)
library(tidyverse)
library(dplyr)
library(stringr)
if(!require(kableExtra)) install.packages("kableExtra")
library(kableExtra)
if(!require(plotly)) install.packages("plotly", repos = "http://cran.us.r-project.org")
library(plotly)
if(!require(plotly)) install.packages("DT")
library(DT)
```

```{r read-data}
#read in the dataset

if (!file.exists("COVID19_Vaccination.csv")){
  library("RSocrata")
  vaccination <- read.socrata(
                 "https://data.cdc.gov/resource/n8mc-b4w4.json",
                  app_token = "KS8vICWuRMDR6QzLnGP7SVO1a",
                  email     = "yatingzeng18@gmail.com"
  )
}

vaccination <- read_csv("COVID19_Vaccination.csv")
```

```{r keep data of CA}

#select only the data of CA
ca_vac <- vaccination[which(vaccination$Location == "CA"), ]
#str(ca_vac)

#reorder the dataset by Demographic_Category and then date
ca_vac <- ca_vac[order( ca_vac[,3], ca_vac[,1] ),]
```

```{r Change the names of the key variables}
#head(ca_vac)
#simplified the variable names
colnames(ca_vac)[3]  <- "cat"
colnames(ca_vac)[5]  <- "dose1"
colnames(ca_vac)[6]  <- "series"
colnames(ca_vac)[9]  <- "dose1_pct"
colnames(ca_vac)[10] <- "series_pct"
colnames(ca_vac)[11] <- "booster_pct"
colnames(ca_vac)[12] <- "secbooster_pct"
```

```{r remove missing values}
#checking the proportion of missing values
#(colMeans(is.na(ca_vac)))*100

vac <- ca_vac[which(ca_vac$cat != "Age_Unknown"), ] 
vac <- vac %>%
  filter(!is.na(vac$dose1),!is.na(vac$census))

#check the missing value again
#(colMeans(is.na(vac)))*100

#learn more about the dataset
#check about the "Demographic_Category", "Dose1_pct",and "series_pct"
#unique(vac$cat)
#summary(vac$dose1_pct)
#summary(vac$series_pct)
```

```{r}
#create new variables about date
vac$Date  <- substr(vac$Date, 0, 10)
vac$year  <- substr(vac$Date, 7, 10)
vac$month <- substr(vac$Date, 0, 2)
vac$day   <- substr(vac$Date, 4, 5)

#sort the data by date 
vac1 <- vac[with(vac, order(year, month, day)), ]

#create a new "date" numeric variable with the time order acceptable for reoder dataset
vac1 <- mutate(vac1, date = paste(year, month, day))
vac1$date <- str_replace_all(vac1$date, fixed(" "), "")

vac1$year  <- as.numeric(vac1$year)
vac1$month <- as.numeric(vac1$month)
vac1$day   <- as.numeric(vac1$day)

```

```{r remove the informarion}
#remove the some information of no interest: booster information; the data of level "ages 65+"
vac1 = subset(vac1, select = -c(Booster_Doses, Second_Booster, booster_pct, secbooster_pct) )
vac1 <- vac1 %>%
  filter(vac1$cat != "Ages_65+_yrs",
               vac1$cat != "Female_Ages_65+_yrs",
               vac1$cat != "Male_Ages_65+_yrs"
               )

#find that there is a unreasonable order for the level 5-11

#rename the level of 5-11 to 05-11
vac1$cat <- str_replace_all(vac1$cat, fixed("Female_Ages_5-11_yrs"), "Female_Ages_05-11_yrs")
vac1$cat <- str_replace_all(vac1$cat, fixed("Male_Ages_5-11_yrs"),   "Male_Ages_05-11_yrs")
vac1$cat <- str_replace_all(vac1$cat, fixed("Ages_5-11_yrs"),        "Ages_05-11_yrs")
vac1$cat <- str_replace_all(vac1$cat, fixed("Female_Ages_2-4_yrs"),  "Female_Ages_02-04_yrs")
vac1$cat <- str_replace_all(vac1$cat, fixed("Male_Ages_2-4_yrs"),    "Male_Ages_02-04_yrs")
vac1$cat <- str_replace_all(vac1$cat, fixed("Ages_2-4_yrs"),         "Ages_02-04_yrs")

```

```{r}
#splitting the data by "cat" level into 4 subset: "age"; "Female_age"; "Male_age"; "sex"
vac1$CAT = substr(vac1$cat, 0, 1)
#build a subset for 
vac_age<- vac1 %>%
  filter(vac1$CAT == "A")
vac_Fage<- vac1 %>%
  filter(vac1$CAT == "F")
vac_Mage<- vac1 %>%
  filter(vac1$CAT == "M")
vac_sex<- vac1 %>%
  filter(vac1$CAT == "S")
```

<br>

# **Supplementary tables**

<br>

## 1. Summary tables tables for at least one dose

```{r summary table for dose1}
# summary table for dose1
vac_age_tab <- vac_age %>% group_by(cat) %>%
                   summarise(
                     Min    = min(dose1_pct, na.rm = T),
                     Q1     = quantile(dose1_pct, 0.25),
                     Median = median(dose1_pct),
                     Q3     = quantile(dose1_pct, 0.75),
                     Max    = max(dose1_pct, na.rm = T), 
                     Records    = sum(!is.na(dose1_pct))
                     ) %>%  arrange(cat) %>% 
                              mutate_at(vars(Q1,Median,Q3,Max,Records), funs(round(.,2))) %>% 
                                rename(Group = cat)

vac_Fage_tab <- vac_Fage %>% group_by(cat) %>%
                   summarise(
                     Min    = min(dose1_pct, na.rm = T),
                     Q1     = quantile(dose1_pct, 0.25),
                     Median = median(dose1_pct),
                     Q3     = quantile(dose1_pct, 0.75),
                     Max    = max(dose1_pct, na.rm = T), 
                     Records    = sum(!is.na(dose1_pct))
                     ) %>%  arrange(cat) %>% 
                              mutate_at(vars(Q1,Median,Q3,Max,Records), funs(round(.,2))) %>% 
                                rename(Group = cat)

vac_Mage_tab <- vac_Mage %>% group_by(cat) %>%
                   summarise(
                     Min    = min(dose1_pct, na.rm = T),
                     Q1     = quantile(dose1_pct, 0.25),
                     Median = median(dose1_pct),
                     Q3     = quantile(dose1_pct, 0.75),
                     Max    = max(dose1_pct, na.rm = T), 
                     Records    = sum(!is.na(dose1_pct))
                     ) %>%  arrange(cat) %>% 
                              mutate_at(vars(Q1,Median,Q3,Max,Records), funs(round(.,2))) %>% 
                                rename(Group = cat)

vac_sex_tab <- vac_sex %>% group_by(cat) %>%
                   summarise(
                     Min    = min(dose1_pct, na.rm = T),
                     Q1     = quantile(dose1_pct, 0.25),
                     Median = median(dose1_pct),
                     Q3     = quantile(dose1_pct, 0.75),
                     Max    = max(dose1_pct, na.rm = T), 
                     Records    = sum(!is.na(dose1_pct))
                     ) %>%  arrange(cat) %>% 
                              mutate_at(vars(Q1,Median,Q3,Max,Records), funs(round(.,2))) %>% 
                                rename(Group = cat)
```
<br>

### {.tabset}

#### Supplementary table 1.Summary of Percent of people with at least one dose in females grouped by age

```{r}
datatable(vac_Fage_tab)
```

#### Supplementary table 2.Summary of Percent of people with at least one dose in males grouped by age

```{r}
datatable(vac_Mage_tab)
```

#### Supplementary table 3.Summary of Percent of people with at least one dose grouped by sex

```{r}
datatable(vac_sex_tab)
```

### {-}

## 2. Summary tables for series dose
```{r summary table for series}
# summary table for series dose
vac_age_tab2 <- vac_age %>% group_by(cat) %>%
                   summarise(
                     Min    = min(series_pct, na.rm = T),
                     Q1     = quantile(series_pct, 0.25, na.rm = T),
                     Median = median(series_pct, na.rm = T),
                     Q3     = quantile(series_pct, 0.75, na.rm = T),
                     Max    = max(series_pct, na.rm = T), 
                     Records    = sum(!is.na(series_pct))
                     ) %>%  arrange(cat) %>% 
                              mutate_at(vars(Q1,Median,Q3,Max,Records), funs(round(.,2))) %>% 
                                rename(Group = cat)

vac_Fage_tab2 <- vac_Fage %>% group_by(cat) %>%
                   summarise(
                     Min    = min(series_pct, na.rm = T),
                     Q1     = quantile(series_pct, 0.25, na.rm = T),
                     Median = median(series_pct, na.rm = T),
                     Q3     = quantile(series_pct, 0.75, na.rm = T),
                     Max    = max(series_pct, na.rm = T), 
                     Records    = sum(!is.na(series_pct))
                     ) %>%  arrange(cat) %>% 
                              mutate_at(vars(Q1,Median,Q3,Max,Records), funs(round(.,2))) %>% 
                                rename(Group = cat)

vac_Mage_tab2 <- vac_Mage %>% group_by(cat) %>%
                   summarise(
                     Min    = min(series_pct, na.rm = T),
                     Q1     = quantile(series_pct, 0.25, na.rm = T),
                     Median = median(series_pct, na.rm = T),
                     Q3     = quantile(series_pct, 0.75, na.rm = T),
                     Max    = max(series_pct, na.rm = T), 
                     Records    = sum(!is.na(series_pct))
                     ) %>%  arrange(cat) %>% 
                              mutate_at(vars(Q1,Median,Q3,Max,Records), funs(round(.,2))) %>% 
                                rename(Group = cat)

vac_sex_tab2 <- vac_sex %>% group_by(cat) %>%
                   summarise(
                     Min    = min(series_pct, na.rm = T),
                     Q1     = quantile(series_pct, 0.25, na.rm = T),
                     Median = median(series_pct, na.rm = T),
                     Q3     = quantile(series_pct, 0.75, na.rm = T),
                     Max    = max(series_pct, na.rm = T), 
                     Records    = sum(!is.na(series_pct))
                     ) %>%  arrange(cat) %>% 
                              mutate_at(vars(Q1,Median,Q3,Max,Records), funs(round(.,2))) %>% 
                                rename(Group = cat)
```

### {.tabset}


#### Supplementary table 4.Summary of Percent of people completed a primary series in females grouped by age

```{r}
datatable(vac_Fage_tab2)
```

#### Supplementary table 5.Summary of Percent of people completed a primary series in males grouped by age

```{r}
datatable(vac_Mage_tab2)
```

#### Supplementary table 6.Summary of Percent of people completed a primary series grouped by sex

```{r}
datatable(vac_sex_tab2)
```

### {-}

<br>

# **Supplementary figures**

<br>

## 1. Summary figures for one dose

```{r summary graphs for dose1, fig.height = 5, fig.width = 8}
#summary graphs for dose1
Fage1 <- vac_Fage %>%
    ggplot(aes(x=date, y=dose1_pct)) +
    geom_boxplot(mapping = aes(x = cat, y = dose1_pct, fill = cat)) +
    theme(axis.text.x = element_text(angle = 60, hjust = 1),
          plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
    labs(x = "Age group", y = "Percent of people with at least one dose", fill = 'Age group') 


Mage1 <- vac_Mage %>%
    ggplot(aes(x=date, y=dose1_pct)) +
    geom_boxplot(mapping = aes(x = cat, y = dose1_pct, fill = cat)) +
    theme(axis.text.x = element_text(angle = 60, hjust = 1),
          plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
    labs(x = "Age group", y = "Percent of people with at least one dose", fill = 'Age group')


sex1 <- vac_sex %>%
    ggplot(aes(x=date, y=dose1_pct)) +
    geom_boxplot(mapping = aes(x = cat, y = dose1_pct, fill = cat)) +
    theme(plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
    labs(x = "Sex", y = "Percent of people with at least one dose", fill='Sex group')
```
### {.tabset}

#### Supplementary figure 7.Percent of people with at least one dose of females grouped by age

```{r}
ggplotly(Fage1)
```

#### Supplementary figure 8.Percent of people with at least one dose of males grouped by age

```{r}
ggplotly(Mage1)
```

#### Supplementary figure 9.Percent of people with at least one dose grouped by sex

```{r}
ggplotly(sex1)
```

### {-}

## 2. Summary figures for series dose

```{r summary graphs for series doses, fig.height = 5, fig.width = 9}
#summary graphs for series doses
age2 <- vac_age %>%
    ggplot(aes(x=date, y=series_pct)) +
    geom_boxplot(mapping = aes(x = cat, y = dose1_pct, fill = cat)) +
    theme(axis.text.x = element_text(angle = 60, hjust = 1),
          plot.caption = element_text(hjust = 0.5, size=rel(1.2))) +
    labs(x = "Age group", y = "Percent of people completed a primary series", fill = 'Age group')


Fage2 <- vac_Fage %>%
    ggplot(aes(x=date, y=series_pct)) +
    geom_boxplot(mapping = aes(x = cat, y = dose1_pct, fill = cat)) +
    theme(axis.text.x = element_text(angle = 60, hjust = 1),
          plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
    labs(x = "Age group", y = "Percent of people completed a primary series", fill = 'Age group')

Mage2 <- vac_Mage %>%
    ggplot(aes(x=date, y=series_pct)) +
    geom_boxplot(mapping = aes(x = cat, y = dose1_pct, fill = cat)) +
    theme(axis.text.x = element_text(angle = 60, hjust = 1),
          plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
    labs(x = "Age group", y = "Percent of people completed a primary series", fill = 'Age group')

sex2 <- vac_sex %>%
    ggplot(aes(x=date, y=series_pct)) +
    geom_boxplot(mapping = aes(x = cat, y = dose1_pct, fill = cat)) +
    theme(plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
    labs(x = "Sex", y = "Percent of people completed a primary seriese", fill = 'Sex group')

sex02 <- vac_sex %>%
    ggplot(aes(x = date, y = dose1_pct)) + 
    geom_point(mapping = aes(x = date, y = dose1_pct, color = cat)) +
    scale_x_discrete(breaks=
          c("20201213","20210601","20211201","20220601","20221019")) +
    theme(plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
    labs(x = "Date(yyyymmdd)", y = "Percent of people with at least one dose", col="Sex group")
```
### {.tabset}

#### Supplementary figure 9.Percent of people completed a primary series of males grouped by age

```{r}
ggplotly(Fage2)
```

#### Supplementary figure 10.Percent of people completed a primary series of males grouped by age

```{r}
ggplotly(Mage2)
```

#### Supplementary figure 11.Percent of people completed a primary seriese grouped by sex in boxplot

```{r}
ggplotly(sex2)
```

#### Supplementary figure 12.Percent of people completed a primary seriese grouped by sex in line plot

```{r}
ggplotly(sex02)
```

### {-}

<br>

# **Visualization of the association**

## 1. Visualization for one dose

```{r visualization for first dose,  fig.height = 6, fig.width = 10}
#visualization for first dose
age3 <- vac_age %>%
    ggplot(aes(x = date, y = dose1_pct)) + 
    geom_point(mapping = aes(x = date, y = dose1_pct, color = cat)) +
    scale_x_discrete(breaks=
          c("20201213","20210601","20211201","20220601","20221019")) +
    theme(plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
    labs(x = "Date(yyyymmdd)", y = "Percent of people with at least one dose", col="Age group")
```
### {.tabset}

#### Supplementary figure 13.Percent of people with at least one dose grouped by age

```{r}
ggplotly(age3)
```

### {-}

## 2. Visualization for series dose

```{r visualization for series dose, fig.height = 6, fig.width = 10}
#visualization for series dose
age4 <- vac_age %>%
    ggplot(aes(x=date, y=series_pct)) + 
    geom_point(mapping = aes(x = date, y = dose1_pct, color = cat))  +
    theme(plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
    scale_x_discrete(breaks=          c("20201213","20210601","20211201","20220601","20221019")) +
    labs(x = "Date(yyyymmdd)", y = "Percent of people completed a primary series", col="Age group")
 
sex4 <- vac_sex %>%
    ggplot(aes(x=date, y=series_pct)) + 
    geom_point(mapping = aes(x = date, y = dose1_pct, color = cat))  +
    theme(plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
    scale_x_discrete(breaks=
          c("20201213","20210601","20211201","20220601","20221019")) +
    labs(x = "Date(yyyymmdd)", y = "Percent of people completed a primary series", col="Sex group")

Fage4 <- vac_Fage %>%
    ggplot(aes(x=date, y=series_pct)) + 
    geom_point(mapping = aes(x = date, y = dose1_pct, color = cat))  +
    theme(plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
    scale_x_discrete(breaks=
          c("20201213","20210601","20211201","20220601","20221019")) +
    labs(x = "Date(yyyymmdd)", y = "Percent of people completed a primary series", col="Age group")

Mage4 <- vac_Mage %>%
    ggplot(aes(x=date, y=series_pct)) + 
    geom_point(mapping = aes(x = date, y = dose1_pct, color = cat))  +
    theme(plot.caption = element_text(hjust=0.5, size=rel(1.2))) +
    scale_x_discrete(breaks=
          c("20201213","20210601","20211201","20220601","20221019")) +
    labs(x = "Date(yyyymmdd)", y = "Percent of people completed a primary series", col="Age group") +
    guides(fill=guide_legend(title="Age group"))
```
### {.tabset}


#### Supplementary figure 14.Percent of people completed a primary series grouped by age

```{r}
ggplotly(age4)
```

#### Supplementary figure 15.Percent of people completed a primary series grouped by sex

```{r}
ggplotly(Fage4)
```

#### Supplementary figure 16.Percent of people completed a primary series of females grouped by age

```{r}
ggplotly(Fage4)
```

#### Supplementary figure 17.Percent of people completed a primary series of males grouped by age

```{r}
ggplotly(Mage4)
```

### {-}


<br>